Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2017-09-13 writeup/conStruct.tex
minor edits

An additional heuristic to evaluate the inclusion of additional layers is to calculate 
each layer's relative contribution to total covariance as their importance, 
and to choose a value of $K$ where all layers have an importance above some cutoff (e.g. 0.1\%).
Because the prior on admixture proportions is a Dirichlet distribution 
that is quite harsh against intermediate admixture values (see Table \ref{tab:param_prior_tab}),  
the model will tend to ``not use" unnecessary layers, 
and they will have a low contribution to overall covariance.
The contribution of the $k$th layer can be calculated as
its contribution to the parametric covariance between samples 
summed over the upper triangle (excluding the diagonal) of the covariance matrix.
The importance of the $k$th layer is its contribution relative to those of all other layers, 
i.e., its contribution divided by the sum of the contributions of all $K$ layers
(for details on calculating this importance, see Appendix \secref{layer_importance}). 
This layer importance is similar to the ``ancestry contribution'' proposed by \cite{fastStructure}. 
However, each of our layers can induce a different amount of covariance between samples embedded in them, 
so we take that into account when calculating each layer's contribution to the whole.

@@

Finally, we calculated layer importance for 
analyses run in each simulation 
(Fig \ref{simK1_laycon}, \ref{simK2_laycon}, and \ref{simK3_laycon}).
In the spatial analyses, once we pass the true $K$, 
subsequent layers add little in terms of (co)variance explained; 
in contrast, additional layers in the nonspatial analyses continue to
be important for describing variation, 
as they are forced to partition up the spatial patterns within clusters.

@@

\begin{figure}
	\centering
		{\includegraphics[width=5in,height=2.5in]{figs/sims/simK1_laycon_barplots.pdf}}
		\caption{
			Layer ``importance" (i.e., how much total covariance is contributed by each layer), 
			for all layers estimated in runs using $K = 1$ through 7 
			for the spatial model (left) and the nonspatial model (right).
			Data were simulated using $K=1$.
			For each value of $K$ along the x-axis, there are an equal number of layer contributions plotted.
		}\label{simK1_laycon}
\end{figure}

@@

\begin{figure}
	\centering
		{\includegraphics[width=6in,height=3in]{figs/bears/bears_laycon_barplots.pdf}}
	\caption{
	Layer ``contributions" (i.e., how much total covariance is contributed by each layer), 
	for all layers estimated in runs using $K = 1$ through 7 
	for the spatial model (left) and the nonspatial model (right).
	For each value of $K$ along the x-axis, there are an equal number of layer contributions plotted.
	Colors are consistent with Fig \ref{bear_K3}.
    }\label{bear_laycon}
\end{figure}

Bears in the sample are partitioned into two main groups ($K \geq 2$):
one (red) to the east of the Rocky Mountains, 
which also occurs in Alaska,
the other primarily west of the Rockies (blue).
The disjointed range of the blue layer likely reflects the fact
that Canada was not sampled, 
and so the blue layer may extend
through the intervening (unsampled) northern Great Plains and Canadian Shield, 
with the red layer presumably then stretching up into British Columbia.

Additional layers in the spatial model have strong statistical support 
up until around $K=5$ or 6 (Fig \ref{bear_xvals}).
However, upon examining layer importance (Fig \ref{bear_laycon}),
it is clear that the additional spatial layers beyond $K=2$ are contributing only a very small amount toward 
total covariance between individuals.
The locations of the individuals inferred to be admixed between the two layers 
provide information about a possible zone of contact and migration between these groups 
along the northwestern coast of North America and near the Cascade Range in the northern U.S.
These findings are consistent with a scenario of postglacial expansion from two refugia, 
one in the American Southwest and one in the American Southeast, 
but lack of any samples from Canada and Mexico, 
and lack of denser sampling across northern North America, 
make more detailed interpretations untrustworthy.
%In addition, our model is not an explicit test of any demographic scenario, 
%and also uses a dataset that consists solely of nuclear (rather than
%mitochondrial) DNA. WHY DO WE CARE about BEAR MTDNS

@@

\paragraph{Comparison to nonspatial model-based clustering}
Above, we showed that 
(a) the nonspatial \texttt{conStruct} model recapitulates results of 
other, commonly-used nonspatial clustering methods,
and 
(b) \texttt{conStruct} can concisely capture spatial structure, which
will be common within populations.
Given this, when should methods without spatial capability be used?
One advantage these have over \texttt{conStruct} is speed.
Although \texttt{conStruct}'s computation time is invariant 
with the number of loci included in the dataset 
(after the initial calculation of the allelic covariance), 
it currently scales poorly with number of samples.
The computationally limiting step is the inversion of the parametric covariance, 
which is an operation of order $O(n^3)$, 
whereas computation time for, e.g., STRUCTURE, 
scales linearly with number of samples.
Our speed, on datasets with large sample sizes, 
could be improved by adopting rank-one updates 
to the inverse of the covariance matrix 
\citep[e.g.][]{woodbury1950,sherman_morrison1950}
when updating a sample's admixture proportions, 
which alters only a single row/column of the covariance matrix. 
However, we have not implemented this yet, 
as it would likely mean losing the ability to do efficient,  
Hamiltonian Monte Carlo sampling of our parameters.

\gc{SO I THINK IF YOU SAY THIS you need to give some run time specs,
  mainly to show that it's not insanely slow.}

@@

From a biological perspective, 
it is important to stress that patterns of relatedness between individuals and populations 
are shaped by complex spatial and hierarchical processes.
All individuals within a species (and indeed, all individuals across all species), 
are related to one another in some way, 
and summarizing that relatedness with a single value of $K$ may be reductive or misleading.
We therefore encourage users to perform analyses across different values of $K$ and 
observe which layers split out at what levels 
(this is conceptually similar to taking successively shallower
cross-sections of the population phylogeny), 
and also to take the results of the proposed cross-validation procedure with a large grain of salt.
Calculating layer importance may also be a useful heuristic, 
as it can reveal layers with statistical support but small biological import.

@@

\paragraph{Implications for management and conservation}
If isolation by distance is common, 
then a likely result of applying \texttt{conStruct} to existing data is that 
populations previously identified using nonspatial clustering methods 
will be collapsed into each other.  
This ``lumping" might better reflect biological ``reality'',  
but may also have implications for management decisions and conservation policy, 
both of which are often predicated on the identification of discrete ``management units" (MUs) 
identified using genetic data \citep{Moritz1994,Waples_1998,Moritz_etal_2002}.

It is therefore important to stress that individuals sampled from the same \texttt{conStruct} layer 
may be quite genetically diverged from one another, 
perhaps especially at loci underlying adaptive traits, 
and that a \texttt{conStruct} layer may still contain multiple distinct MUs.  
Alternatively, the collapse of multiple, previously identified MUs into a single \texttt{conStruct} layer 
may reflect a biological reality that these populations are currently 
(or were recently) exchanging migrants, 
and thus might emphasize the importance of maintaining habitat corridors between demes, 
or of implementing an integrated conservation plan across multiple demes within a layer.

@@

\newpage
\section{Calculating layer importance}\label{layer_importance}
First, we describe the allelic covariance between two samples, $i$, and $j$.
Note that this is the allelic covariance described in Eqn \eqref{allelic_covariance},
in which we randomly polarize alleles,
and not the genetic covariance, which subtracts mean allele frequencies.
Let $A$ and $B$ be a randomly chosen alleles from the two samples, respectively,
at a randomly chosen locus 
(if the two populations are the same, then choose without replacement).
Then, if we let $X=2(A-1/2)$ and $Y=2(B-1/2)$,
since $X$ and $Y$ take the values $\pm 1$,
and thanks to the random polarization, $\E[A] = \E[A] = 1/2$,
$$\begin{aligned}
\E[XY]
        &= \P\{ X=Y \} - \P\{ X \neq Y \} \\
        &= 2 \P\{ X=Y \} - 1 \\
        &= 1 - 2 \P\{ X \neq Y \} .
\end{aligned}$$
To translate, $\P\{ X=Y \}$ is the probability that the alleles from our two focal samples agree with each other,
while $\P\{ X \neq Y \}$ is the probability that they disagree.
For the first thing to happen, $i$ and $j$ must coalesce before coalescing with the reference samples,
so for $X$ and $Y$ to agree they must be more closely related than typical in the population.
The last part, $\P\{ XY \neq 0\}$, is the probability that both samples disagree from the references.
(This implies that $\E[XY] = 1 - 2 D_{ij}$, 
where $D_{ij}$ is the probability that two randomly chosen alleles differ, 
which is the genetic divergence.)

@@

%
\begin{equation}
\xi^{(k)} = 
	\sum\limits_{i=1}^{N-1}
		\sum\limits_{j=i+1}^N
			w^{(k)}_iw^{(k)}_j 
				\left(	
					2\alpha^{(k)}_0 
						\left( \text{exp} 
							\left( -(\alpha^{(k)}_D D_{i,j}) ^ {\alpha^{(k)}_2}\right) 
						\right) + 2\phi^{(k)} + \frac{1}{2}
				\right)
\label{layer_contribution}
\end{equation}
%
which is that layer's contribution to agreement between samples 
summed over the upper triangle (excluding the diagonal) of the covariance matrix.
We define the importance of the $k$th layer, $\Xi^{(k)}$, 
as the relative contribution of the $k$th layer to total agreement:
%
\begin{equation}
\Xi^{(k)} = 
\frac{\xi^{(k)}}
	{\sum\limits_{k=1}^{K}
	    	\xi^{(k)}}
\label{rel_layer_contribution}
\end{equation}
%

@@

\begin{figure}
	\centering
		{\includegraphics[width=5in,height=2.5in]{figs/sims/simK2_laycon_barplots.pdf}}
		\caption{
			Layer ``importance" (i.e., how much total covariance is contributed by each layer), 
			for all layers estimated in runs using $K = 1$ through 7 
			for the spatial model (left) and the nonspatial model (right).
			Data were simulated using $K=2$.
			For each value of $K$ along the x-axis, there are an equal number of layer contributions plotted.
		}\label{simK2_laycon}
\end{figure}

@@

\begin{figure}
	\centering
		{\includegraphics[width=5in,height=2.5in]{figs/sims/simK3_laycon_barplots.pdf}}
		\caption{
			Layer ``importance" (i.e., how much total covariance is contributed by each layer), 
			for all layers estimated in runs using $K = 1$ through 7 
			for the spatial model (left) and the nonspatial model (right).
			Data were simulated using $K=3$.
			For each value of $K$ along the x-axis, there are an equal number of layer contributions plotted.
		}\label{simK3_laycon}
\end{figure}

@@

\begin{figure}
	\centering
		{\includegraphics[width=6in,height=3in]{figs/populus/populus_laycon_barplots.pdf}}
	\caption{
	Layer ``contributions" (i.e., how much total covariance is contributed by each layer), 
	for all layers estimated in runs using $K = 1$ through 7 
	for the spatial model (left) and the nonspatial model (right).
	For each value of $K$ along the x-axis, there are an equal number of layer contributions plotted.
	Colors are consistent with Figs \ref{populus_sp_pies}, \ref{populus_sp_clst_covs}, \ref{populus_nsp_pies}, and \ref{populus_nsp_clst_covs}.
    }\label{populus_laycon}
\end{figure}

>>>>

An additional heuristic to evaluate the inclusion of additional layers is to calculate 
each layer's relative contribution to total covariance, 
and to choose a value of $K$ where all layers have a contribution above some cutoff (e.g. 0.1\%).
Because the prior on admixture proportions is a Dirichlet distribution 
that is quite harsh against intermediate admixture values (see Table \ref{tab:param_prior_tab}),  
the model will tend to ``not use" unnecessary layers, 
and they will have a low contribution to overall covariance.
The contribution of the $k$th layer can be calculated as
its contribution to the parametric covariance between samples 
summed over the upper triangle (excluding the diagonal) of the covariance matrix.
The contribution of the $k$th layer is its absolute contribution relative to those of all other layers, 
i.e., its contribution divided by the sum of the contributions of all $K$ layers
(for details on calculating this contribution, see Appendix \secref{layer_contribution}). 
This layer contribution is similar to the ``ancestry contribution'' proposed by \cite{fastStructure}. 
However, each of our layers can induce a different amount of covariance between samples embedded in them, 
so we take that into account when calculating each layer's contribution to the whole.

@@

Finally, we calculated layer contribution for 
analyses run in each simulation 
(Fig \ref{simK1_laycon}, \ref{simK2_laycon}, and \ref{simK3_laycon}).
In the spatial analyses, once we pass the true $K$, 
subsequent layers add little in terms of (co)variance explained; 
in contrast, additional layers in the nonspatial analyses continue to
be important for describing variation, 
as they are forced to partition up the spatial patterns within clusters.

@@

\begin{figure}
	\centering
		{\includegraphics[width=5in,height=2.5in]{figs/sims/simK1_laycon_barplots.pdf}}
		\caption{
			Layer contributions (i.e., how much total covariance is contributed by each layer), 
			for all layers estimated in runs using $K = 1$ through 7 
			for the spatial model (left) and the nonspatial model (right).
			Data were simulated using $K=1$.
			For each value of $K$ along the x-axis, there are an equal number of layer contributions plotted.
		}\label{simK1_laycon}
\end{figure}

@@

\begin{figure}
	\centering
		{\includegraphics[width=6in,height=3in]{figs/bears/bears_laycon_barplots.pdf}}
	\caption{
	Layer contributions (i.e., how much total covariance is contributed by each layer), 
	for all layers estimated in runs using $K = 1$ through 7 
	for the spatial model (left) and the nonspatial model (right).
	For each value of $K$ along the x-axis, there are an equal number of layer contributions plotted.
	Colors are consistent with Fig \ref{bear_K3}.
    }\label{bear_laycon}
\end{figure}

Bears in the sample are partitioned into two main groups ($K \geq 2$):
one (red) to the east of the Rocky Mountains, 
which also occurs in Alaska,
the other primarily west of the Rockies (blue).
The disjointed range of the blue layer likely reflects the fact
that Canada was not sampled, 
and so the red layer may extend
through the intervening (unsampled) northern Great Plains and Canadian Shield, 
with the blue layer presumably then stretching up into British Columbia.

Additional layers in the spatial model have strong statistical support 
up until around $K=5$ or 6 (Fig \ref{bear_xvals}).
However, upon examining layer contributions (Fig \ref{bear_laycon}),
it is clear that the additional spatial layers beyond $K=2$ are contributing only a very small amount toward 
total covariance between individuals.
The locations of the individuals inferred to be admixed between the two layers 
provide information about a possible zone of contact and migration between these groups 
along the northwestern coast of North America and near the Cascade Range in the northern U.S.
These findings are consistent with a scenario of postglacial expansion from two refugia, 
one in the American Southwest and one in the American Southeast, 
but lack of any samples from Canada and Mexico, 
and lack of denser sampling across northern North America, 
make more detailed interpretations untrustworthy.
%In addition, our model is not an explicit test of any demographic scenario, 
%and also uses a dataset that consists solely of nuclear (rather than
%mitochondrial) DNA. WHY DO WE CARE about BEAR MTDNS

@@

\paragraph{Comparison to nonspatial model-based clustering}
Above, we showed that 
(a) the nonspatial \texttt{conStruct} model recapitulates results of 
other, commonly-used nonspatial clustering methods,
and 
(b) \texttt{conStruct} can concisely capture spatial structure, which
will be common within populations.
Given this, when should methods without spatial capability be used?
One advantage these have over \texttt{conStruct} is speed.
Although \texttt{conStruct}'s computation time is invariant 
with the number of loci included in the dataset 
(after the initial calculation of the allelic covariance), 
it currently scales poorly with number of samples.
The computationally limiting step is the inversion of the parametric covariance, 
which is an operation of order $O(n^3)$, 
whereas computation time for, e.g., STRUCTURE, 
scales linearly with number of samples.
Our speed, on datasets with large sample sizes, 
could be improved by adopting rank-one updates 
to the inverse of the covariance matrix 
\citep[e.g.][]{woodbury1950,sherman_morrison1950}
when updating a sample's admixture proportions, 
which alters only a single row/column of the covariance matrix. 
However, we have not implemented this yet, 
as it would likely mean losing the ability to do efficient,  
Hamiltonian Monte Carlo sampling of our parameters.
On a desktop machine with a 4.2 GHz Intel Core i7 processor, 
an analysis of the the poplar dataset (35 samples, 22,500 loci), 
running the spatial model with 7 layers for 5,000 MCMC iterations 
took 29.8 minutes.
For the black bear dataset (84 samples, 21,000 loci) 
running the spatial model with 7 layers for 5,000 MCMC iterations, 
which was more than sufficient for convergence,
took 6.9 hours.

@@

From a biological perspective, 
it is important to stress that patterns of relatedness between individuals and populations 
are shaped by complex spatial and hierarchical processes.
All individuals within a species (and indeed, all individuals across all species), 
are related to one another in some way, 
and summarizing that relatedness with a single value of $K$ may be reductive or misleading.
We therefore encourage users to perform analyses across different values of $K$ and 
observe which layers split out at what levels 
(this is conceptually similar to taking successively shallower
cross-sections of the population phylogeny), 
and also to take the results of the proposed cross-validation procedure with a large grain of salt.
Calculating layer contributions may also be a useful heuristic, 
as it can reveal layers with statistical support but small biological import.

@@

\paragraph{Implications for management and conservation}
If isolation by distance is common, 
then a likely result of applying \texttt{conStruct} to existing data is that 
populations previously identified using nonspatial clustering methods 
will be collapsed into each other.  
This ``lumping" might better reflect biological reality, 
but may also have implications for management decisions and conservation policy, 
both of which are often predicated on the identification of discrete ``management units" (MUs) 
identified using genetic data \citep{Moritz1994,Waples_1998,Moritz_etal_2002}.

It is therefore important to stress that individuals sampled from the same \texttt{conStruct} layer 
may be quite genetically diverged from one another, 
perhaps especially at loci underlying adaptive traits, 
and that a \texttt{conStruct} layer may still contain multiple distinct MUs worthy of independent protections.  
Alternatively, the inclusion of multiple MUs into a single \texttt{conStruct} layer 
may occur if these populations are currently 
(or were recently) exchanging migrants, 
and thus might emphasize the importance of maintaining habitat corridors between demes, 
or of implementing an integrated conservation plan across multiple demes within a layer.

@@

\newpage
\section{Calculating layer contributions}\label{layer_contribution}
First, we describe the allelic covariance between two samples, $i$, and $j$.
Note that this is the allelic covariance described in Eqn \eqref{allelic_covariance},
in which we randomly polarize alleles,
and not the genetic covariance, which subtracts mean allele frequencies.
Let $A$ and $B$ be a randomly chosen alleles from the two samples, respectively,
at a randomly chosen locus 
(if the two populations are the same, then choose without replacement).
Then, if we let $X=2(A-1/2)$ and $Y=2(B-1/2)$,
since $X$ and $Y$ take the values $\pm 1$,
and thanks to the random polarization, $\E[A] = \E[A] = 1/2$,
$$\begin{aligned}
\E[XY]
        &= \P\{ X=Y \} - \P\{ X \neq Y \} \\
        &= 2 \P\{ X=Y \} - 1 \\
        &= 1 - 2 \P\{ X \neq Y \} .
\end{aligned}$$
To translate, $\P\{ X=Y \}$ is the probability that the alleles from our two focal samples agree with each other,
while $\P\{ X \neq Y \}$ is the probability that they disagree.
For the first thing to happen, $i$ and $j$ must coalesce before coalescing with the reference samples,
so for $X$ and $Y$ to agree they must be more closely related than typical in the population.
The last part, $\P\{ XY \neq 0\}$, is the probability that both samples disagree from the references.
(This implies that $\E[XY] = 1 - 2 D_{ij}$, 
where $D_{ij}$ is the probability that two randomly chosen alleles differ, 
which is the genetic divergence.)

@@

%
\begin{equation}
\xi^{(k)} = 
	\sum\limits_{i=1}^{N-1}
		\sum\limits_{j=i+1}^N
			w^{(k)}_iw^{(k)}_j 
				\left(	
					2\alpha^{(k)}_0 
						\left( \text{exp} 
							\left( -(\alpha^{(k)}_D D_{i,j}) ^ {\alpha^{(k)}_2}\right) 
						\right) + 2\phi^{(k)} + \frac{1}{2}
				\right)
\label{layer_contribution}
\end{equation}
%
which is that layer's contribution to agreement between samples 
summed over the upper triangle (excluding the diagonal) of the covariance matrix.
We define the contribution of the $k$th layer, $\Xi^{(k)}$, 
as the relative contribution of the $k$th layer to total agreement:
%
\begin{equation}
\Xi^{(k)} = 
\frac{\xi^{(k)}}
	{\sum\limits_{k=1}^{K}
	    	\xi^{(k)}}
\label{rel_layer_contribution}
\end{equation}
%

@@

\begin{figure}
	\centering
		{\includegraphics[width=5in,height=2.5in]{figs/sims/simK2_laycon_barplots.pdf}}
		\caption{
			Layer contributions (i.e., how much total covariance is contributed by each layer), 
			for all layers estimated in runs using $K = 1$ through 7 
			for the spatial model (left) and the nonspatial model (right).
			Data were simulated using $K=2$.
			For each value of $K$ along the x-axis, there are an equal number of layer contributions plotted.
		}\label{simK2_laycon}
\end{figure}

@@

\begin{figure}
	\centering
		{\includegraphics[width=5in,height=2.5in]{figs/sims/simK3_laycon_barplots.pdf}}
		\caption{
			Layer contributions (i.e., how much total covariance is contributed by each layer), 
			for all layers estimated in runs using $K = 1$ through 7 
			for the spatial model (left) and the nonspatial model (right).
			Data were simulated using $K=3$.
			For each value of $K$ along the x-axis, there are an equal number of layer contributions plotted.
		}\label{simK3_laycon}
\end{figure}

@@

\begin{figure}
	\centering
		{\includegraphics[width=6in,height=3in]{figs/populus/populus_laycon_barplots.pdf}}
	\caption{
	Layer contributions (i.e., how much total covariance is contributed by each layer), 
	for all layers estimated in runs using $K = 1$ through 7 
	for the spatial model (left) and the nonspatial model (right).
	For each value of $K$ along the x-axis, there are an equal number of layer contributions plotted.
	Colors are consistent with Figs \ref{populus_sp_pies}, \ref{populus_sp_clst_covs}, \ref{populus_nsp_pies}, and \ref{populus_nsp_clst_covs}.
    }\label{populus_laycon}
\end{figure}
